Uncertainty in the evolution of social learning

### 2022-05-04 ms.tex
updated main results figure with improved design
updated appendix figs
full outline of analysis, now filling it out

% \documentclass[11pt,letterpaper]{article}
\documentclass[letterpaper,11.5pt]{scrartcl}
% \documentclass[11pt]{report}
% \documentclass{report}
% \documentclass{book}
\usepackage[bookmarks, hidelinks]{hyperref}
\usepackage{amssymb,amsmath}
% \usepackage{fullpage}
\usepackage{tabulary}
\usepackage{tabularx}
\usepackage{float}
\usepackage[margin=1.00in]{geometry}
% \usepackage[margin=0.90in]{geometry}

@@

\title{Some forms of uncertainty may suppress the evolution of social learning}

@@

\begin{figure}
  \caption{Model dynamics are composed of two major parts: first, agents perform
  behaviors for $M$ time steps in a single generation. Then agents reproduce
  randomly, with more successful agents being more likely to reproduce. Newly
  ``born'' social learner agents use success bias to select and learn from a chosen
  teacher. Then, all agents from the previous generation die off, all new agents
  begin a new generation, and the process continues until 1000 total time steps
  have elapsed. Agents are initialized as ``blank slates''.}
  \centering
    \includegraphics[width=0.9\textwidth]{Figures/IntraInterGenerationalDynamics.pdf}
  \label{fig:IntraInterGenerationalDynamics}
\end{figure}

@@

% \begin{itemize}
%   \item 
%     To understand how different forms of uncertainty affect social learning evolution
%     we systematically varied environmental variables in our model and observed
%     corresponding changes in mean social learning prevalence, $\meansl$. We
%     confirmed that, in many cases, increased environmental variability $u$ 
%     suppresses social learning. However\ldots

%   \item 
%     Our analysis confirmed existing theoretical predictions that social learning
%     is valuable only when environments are sufficiently stable, i.e., 
%     environmental variability $u$ is sufficiently small.
%     Because of the central role of $u$, it is set on the x-axis in
%     all analysis plots shown in Figure~\ref{fig:results}. 
%     The threshold over $u$ at which social learning begins to decline
%     varies depending on the value of other environmental parameters.

%   \item 
%     However, more possible behaviors $B$ ($B=2,4,10$ for the first, second, and
%     third column respectively in Figure~\ref{fig:results}) can lead
%     social learning to be selected for even for large values of $u$, possibly
%     due to the low risk and potentially high reward of learning an accurate
%     tally of mean payoffs from the previous generation. 
%     Since there are many behaviors and
%     few observations, softmax weighting will not significantly 
%     favor a sub-optimal behavior. However, if one does learn an accurate tally
%     then one does have an increased chance of performing the best one.

%   \item 
%     $\pilow$ can alter selection selection pressure for or
%     against social learning, depending on the number of time steps per generation.
%     For instance, when $B=2$, increasing $\pilow$ (i.e. increased uncertainty
%     about which behavior is optimal) leads at first to less sharp sigmoid curve, 
%     where social learning is less prevalent with increased $\pilow$ when
%     $u < 0.5$ and social learning is more prevalent with increased $\pilow$
%     when $u > 0.5$. When $\pilow = 0.8$ noise and genetic drift dominate across
%     tested values of $B$, with possibly a slight inverse correlation 
%     in social learning as $u$ increases. When $B \in \{4,10\}$, increasing
%     $\pilow \in \{0.1, 0.5\}$ leads uniformly to increased $\meansl$ over 
%     $u$ in those settings.

%   \item
%     If $M/B$ has an effect it is mostly to 
%     increase reliance on social learning due to social information being of 
%     higher quality since teachers would have had more learning opportunities
%     for larger $M$. (THERE ARE CASES WHERE THIS IS NOT THE CASE BUT THERE IS 
%     STILL TOO MUCH NOISE DUE TO SMALL NTRIALS)

%   \item
%     Synthesis/analysis conclusion

% \end{itemize}

@@

@@

>>>>

% \documentclass[11pt,letterpaper]{article}
\documentclass[letterpaper,11.5pt]{scrartcl}
% \documentclass[11pt]{report}
% \documentclass{report}
% \documentclass{book}
\usepackage[bookmarks, hidelinks]{hyperref}
\usepackage[chatter]{rotating}
\usepackage{amssymb,amsmath}
\usepackage[title]{appendix}
% \usepackage{fullpage}
\usepackage{tabulary}
\usepackage{tabularx}
\usepackage{float}
% \usepackage[margin=0.50in]{geometry}
\usepackage[margin=1.00in]{geometry}

@@

\title{Uncertainty in the evolution of social learning}

@@

\begin{figure}
  \caption{Model dynamics are composed of two major parts: first, agents perform
  behaviors for $M$ time steps in a single generation. Then agents reproduce
  randomly, with more successful agents being more likely to reproduce. Newly
  ``born'' social learner agents use success bias to select and learn from a chosen
  teacher. Then, all agents from the previous generation die off, all new agents
  begin a new generation, and the process continues until 1000 total time steps
  have elapsed. Agents are initialized as ``blank slates''.}
  \label{fig:IntraInterGenerationalDynamics}
  \centering
    \includegraphics[width=0.9\textwidth]{Figures/IntraInterGenerationalDynamics.pdf}
\end{figure}

@@

We analyzed the outcomes of model runs across systematically varied uncertainty
parameters over 1000 trials for each parameter setting. 
We observed complex outcomes where the effects of uncertainty parameters
were highly dependent on one another as measured by the mean prevalence of
social learning, $\meansl$, across trials for each parameter setting. 
The prevalence of social learning
generally decreased with environmental variability, $u$, as expected from standard
models of social learning evolution. However, the shape of the decrease was not
uniform (Figure~\ref{fig:evolutionOfSL}). 
We observed that social learning did in fact enable populations to 
do better than individual learners, but only when (multi-dimensional) uncertainty 
were sufficiently limited (Figure~\ref{fig:meanNetPayoffs}).

\subsection{Evolution of social learning under uncertainty}
The shape of the decrease from all agents
always becoming social learners ($\meansl = 1.0$) to 
all agents always becoming individual learners ($\meansl = 0.0$)
is heterogeneous. We identified patterns in this heterogeneity that inform our
understanding of when and how social learning is advantageous, and when 
and why individual learning is advantageous. 
Changes in the uncertainty parameters ($\pilow$, $B$, $L$) 
result in the following changes in the shape of the transition from $\meansl = 1.0$ 
to $\meansl = 0.0$: (1) the transition
occurs at different critical values of $u$; and (2) the transition occurs
with greater or lesser rapidity over $u$. 
For instance, when $B=2$, $\pilow = 0.1$, and $L=8$, $\meansl$ goes from 1 to 
0 starting at $u=0.1$ and reaches 0 at $u=0.3$ (Figure~\ref{fig:evolutionOfSL},
top left). When we increased $B=4$ (Figure~\ref{fig:evolutionOfSL}, top center), the
transition from $\meansl = 1$ to 0 begins at $u=0.3$. When $B=4$, $\pilow=0.1$,
and $L=1$ the transition from $\meansl=1$ to 0 begins at $u=0.3$ but does not
reach 0 until $u \approx 0.9$ (again Figure~\ref{fig:evolutionOfSL}, top center).

We can draw some conclusions about the evolution of social learning by analyzing the
location and shape of the suppression of social learning as $u$ was increased.
First, let us observe some general trends. Increased $\pilow$ generally
flattened the suppression of social learning as $u$ increased.  
Increased $B$ tended to increase the critical value of $u$ at which the 
system transitions from favoring social learning to favor individual learning.  
Increased $L$ generally sharpened the suppression of social learning.
Within each of these broad trends over the uncertainty parameters are several
details that depend on context as defined by the values of the other uncertainty
parameters.

When $\pilow$ was increased, the suppression of social learning was, in
general, less rapid as $u$ increased. This effect is particularly acute when
$\pilow = 0.8$, which is very nearly equal to the expected value of the optimal
payoff, $\pihigh = 0.9$. When $\pilow = 0.8$ $\meansl$ never actually reaches 
1 or 0 for any parameter settings. The suppression of social learning in this
case is approximately linear, if there is selection at all. When $\pilow = 0.8$
and $L = 1$ there is little to no selection across $B$, 
so populations evolve randomly towards $\meansl = 1$ or $\meansl = 0$.

Increased $B$ generally led to increased critical $u$ at which social learning
starts to be suppresed. When payoffs are not ambiguous ($\pilow = 0.1$), the the
suppression of social learning starts at $u \approx 0.1$ for $B=2$; at $u \approx
0.3$ when $B=4$; and at $u \approx 0.5$ when $B=10$.  When payoffs are more
ambiguous ($\pilow = 0.45,0.8$), increased $B$ also led to a flattening of the
suppression of social learning when $L=1$.

The general effect of $L$ is to increase the rapidity of the suppression of social
learning via the strengthening of selection pressure since longer-lived agents
build up stronger priors over their lifetimes. If it is advantageous to
learn socially, this will be reliably detected. Inversely, decreased $L$
can weaken selection pressure since agents cannot amass useful information when
$L$ is small. This is most extreme when $\pilow = 0.8$ and there is little to
no selection for or against social learning when $L=1$. However, decreased $L$ can
also delay the suppression of social learning as $u$ increases, since social
learning provdes agents with the possibility of learning what the optimal behavior
might be.  Even when $u$ was large, agents with a lifespan of 1 more often evolved
to be social learners because social learning was to be preferred over random
chance.

\subsection{Social learning capacity increases payoffs}

For social learning to evolve, it must improve upon individual learning. 
We found that different parameter settings change how advantageous social learning
is over and above individual learning. To analyze the effect of uncertainty on
payoffs, it is useful to identify some constant reference values which we will
use to evaluate the benefit of social learning. First, the optimal expected payoff
is the optimal behavior payoff times the lifespan, which we can denote
$\langle \pi^*(L) \rangle = L\cdot\pihigh$. The expected base payoff by chance is 
$\langle \pi_{\text{base}} \rangle = L \left[ (B - 1) \pilow + \pihigh \right]$, which
an agent would receive on average from performing purely random behaviors. 
Agents learn adaptively over their lifespan (via the softmax algorithm in 
our model), whether or not they are social learners. We calculated this
value for reference for each triplet of $\pilow$, $B$, and $L$ 
\mt{calculate this and add to plot as a guide}. 
Social learning evolves when the expected payoff of
social learners exceeds the expected payoff of individual learners.

\mt{Use above framework to analyze results after plot guides added as
promised}

\subsection{Evolutionary outcomes vary in their certainty}

Finally, we found that the \emph{evolutionary uncertainty} of the system increases
around the values of $u$ where $\meansl$ goes from 1 to 0. 
In one sense
To confirm and expand our analysis of this uncertainty, we 
also measured environmental uncertainty in terms of the
expected number of time steps to convergence, $\langle T \rangle$. Evolutionary
uncertainty is an outcome measure---an emergent form of uncertainty due to
weak selection pressures. When 
$\meansl \approx 0.5$ this means that approximately equal numbers of trials
ended with all agents as social learners as they ended with all agents 
individual learners. We found that increased $\pilow$
tended to increase the overall uncertainty in terms of $\langle T \rangle$.
Increasing $B$ also increased the $u$ at which $\langle T \rangle$ peaked.
This indicates the evolutionary system is significantly less certain of 
which behavior is optimal around critical uncertainty parameter ensembles.

\mt{interpret this further}

@@

\include{meanNetPayoffs.tex}

\include{stepResultsPlots.tex}

@@

\appendix

\section{Appendix}

\subsection{Convergence information}

\begin{table}[h]
  \caption{Max. iterations = 5000.}
  \centering
  \begin{tabular}{cccc} \toprule
    $B$ & \# not at fixation & \# total series & Pct. not fixated \\
    \midrule  
    2  & 0  & 132000 & 0.0 \% \\
    4  & 0  & 132000 & 0.0 \% \\
    10 & 44 & 132000 & 0.00033  \% \\
    \bottomrule
  \end{tabular} 
\end{table}
